---
title: "Worlds End Traverse Analysis"
author: "Tomas Castillo"
date: "June 30, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("C:/Users/Tomas/OneDrive/Projects/Trail_DataProject/Worlds End Traverse/")
we.traverse <- read.csv("data/WorldsEndTraverse.csv", stringsAsFactors = F)

library(ggplot2)
library(ggthemes)
library(dplyr)
library(scales)
library(gridExtra)
```

## Exploring Personal Data

Last weekend, I took a trip out to Worlds End State Park in the Northeast part of Pennyslvania. It's nestled in the heart of
the Endless Mountains and the Loyalsock State Forest. The route that I picked was not my own choosing,
but is in fact a race course for the Worlds End 100k.

I was treating this run as a training experience and a way to explore some of the beautiful and wild parts of the
state. And that it was.

I was exploring some other blogs on racing and came across someone who did an analysis on their own training efforts that
compared their average paces in each category of grades. I hadn't thought of looking at this approach and decided to pull out
my effort at Worlds End to see how my climbing pace looks at different grades.

I tracked the Traverse with a Suunto Ambit3 Peak, and then pulled the data out from Movescount.

This way, the data looks similar to what would show up on Movescount, Strava, TrainingPeaks, MapMyRun, or whatever else
might display the GPS data.

If we can explore this to see what we've got, we'll first just do an overview of what the climbing profile looks like.

```{r vert profile, echo = F}
ggplot(subset(we.traverse, Altitude > 0), aes(x = miles, y = Altitude)) +
  geom_line(size = 0.5) + 
  theme_hc() +
  labs(x = "Distance [miles]", y = "Altitude [ft]")
```

Some nice climbs and descents, and all of it was technical! Ha... good stuff. I should point out that
the race is a 100k category and comes out to 63 miles. I've got a few excuses for not finishing the full 63 miles, but
a majority of the reason was my knees had taken a beating through all the technicality, the heat had taken its toll on
my overall strength, and it was getting late and I didn't want to run in the dark since I was alone (I do, in fact,
get scared). My total mileage came out to 45 miles. Not bad if you ask me, especially considering the longest
I had run before was 36 miles.

Now that we've got that explanation out of the way, let's look a the paces and how the grade of elevation affects it.

First, we'll just look at the grade of the course.

```{r grade, echo = F}
ggplot(we.traverse, aes(x = miles, y = Grade)) +
  geom_point() +
  theme_hc() +
  labs(x = "Distance [miles]", y = "Grade [%]")
```

Something to notice about the data and then to think about in terms of what it means. There are some points over 100%, and
even some descents over -200%. Recall that grade is vertical gain/horizontal distance, as a percent (so 1% = 1 ft up for 100 ft
horizontal).

Then, if we apply some old fashioned trigonometry, we can figure out the angle. The max grade was
`r format(max(we.traverse$Grade),digits =3)`% giving it an angle (in degrees) of `r format(atan(max(we.traverse$Grade))*180/pi, digits = 3)`.
That's pretty steep, but actually not unreasonable - it's just really steep.

The negative grade of 200% gives us a descent angle of 64. In thinking back, I can probably guess at which point that was
because I was going on very slowly and trying to jump from rock to rock as I was descending.

Alright, so we'll set up some bins that will categorize the grade. The reasoning behind the categories below are what goes from
extremely steep, towards a more runnable grade, to flat. Ideally, the grade would be stepped in 2.5% increments, but because
of the large range, we'll lump some of the more extreme grades together.


```{r Grade Histograme}
ggplot(we.traverse, aes(Grade)) +
  geom_histogram(binwidth = 0.02) +
  coord_trans(y = 'sqrt') +
  scale_y_continuous(minor_breaks = seq(0, 6000, 150),
                     breaks = seq(0, 10000, 2500)) +
  scale_x_continuous(breaks = seq(-2, 150, 0.25),
                     labels = paste0(seq(-2, 150, 0.25)*100, "%")) +
  theme_hc() +
  theme(panel.grid.minor.y = element_line(color = "#d5dbdd", size = 0.5))
```

As one would expect, most of the grades are actually reasonable and below 50% (both positive and negative).
We'll add the bins and then plot accordingly.

```{r Setting Up Grades, message = F}
# Not too many steep grades
gradeBin <- c(-2.5, -1, -0.6, -0.3, -0.25, -0.20, -0.15, -0.125, -0.10, -0.075, -0.05, -0.025, 0,
              0.025, 0.05, 0.075, 0.10, 0.125, 0.15, 0.20, 0.25, 0.3, 0.6, 1, 2.5)
```

```{r time and bins, echo = F, message = F}

# Initialize Grade Rank Column
we.traverse$gradeRank <- 0

# Step through list of grades and categorize
for (i in 1:(length(gradeBin) - 1)) {
  we.traverse <- we.traverse %>% 
    mutate(gradeRank =
             ifelse((gradeRank == 0 & Grade >= gradeBin[i] & Grade < gradeBin[i + 1]),
                    i,
                    gradeRank))
}


## Format the Local Time into time format and add duration in seconds
we.traverse$LocalTime <- as.POSIXct(we.traverse$LocalTime)
startTime <- we.traverse$LocalTime[1]
we.traverse <- we.traverse %>% mutate(Duration = as.numeric(LocalTime - startTime))
```

As I was going through this, I realized that we should take out any points that have a speed of 0. The reason for
that is it's just time that I was still at an aid station (my girlfriend's car), resting, or I was navigating
trying to find my way back on the trail. Since my watch also records `Cadence`, we can take out any points
I was standing still with a `Cadence` of 0 as well.

Because we're only interested in the metrics related to moving, we can subset the data.

```{r eliminate stops}
we.traverse <- subset(we.traverse, Speed > 0 & Cadence > 0)
```


```{r grade speed}
ggplot(we.traverse, aes(x = gradeRank, y = speed.km)) + 
  geom_point(position = position_jitter(width = 1), alpha = 0.2, col = '#b0bcbf') +
  geom_point(stat = 'summary', fun.y = mean, col = '#1b5b83', size =2) +
  scale_x_continuous("Grade Bins",
                     breaks = seq(1, 25,1),
                     labels = paste0(format(gradeBin*100, digits = 0),"%")) +
  scale_y_continuous("Speed [km/h]") +
  theme_hc() +
  geom_smooth(method = 'lm', formula = y ~ x*I(x^2)) +
  theme(axis.text.x = element_text(size = 8,
                                   angle = 90, vjust = 0.5),
        panel.grid.major.x = element_line(color = "light grey", 
                                          linetype = 2))
```

While this fit seems to work, I don't usually look at my speed, but rather at my pace. So let's convert
`Speed` to `pace`. I've got it in _km/h_ so we'll need to do a few mathematical transformations first.

```{r adding pace, message = FALSE}
we.traverse <- we.traverse %>% 
  mutate(pace = (speed.km*(0.621/60))^(-1))

```

We'll do the same plot as above, but this time, with pace on the y-axis instead of speed.

```{r grade pace, warning=F}
ggplot(we.traverse, aes(x = gradeRank, y = pace)) + 
  geom_point(position = position_jitter(width = 1, height = 3), alpha = 0.2, col = '#b0bcbf') +
  geom_point(stat = 'summary', fun.y = mean, col = '#1b5b83', size =2) +
  scale_x_continuous("Grade",
                     breaks = seq(1, 25,1),
                   labels = paste0(format(gradeBin*100, digits = 0),"%")) +
  scale_y_continuous("Pace [min/mile]", breaks = seq(0,75,5), 
                     labels = seq(0,75,5), limits = c(0,40)) +
  theme_hc() +
  geom_smooth(method = 'lm', formula = y ~ x*I(x^2)) +
  theme(axis.text.x = element_text(size = 8,
                                   angle = 90, vjust = 0.5),
        panel.grid.major.x = element_line(color = "light grey", 
                                          linetype = 2))
```

Good stuff.

So what does a linear model look like for pace over grade? What if we throw in some of the other variables?

Let's start with just `pace ~ grade rank`.

```{r pace grade rank}
pg <- lm(pace ~ gradeRank, data = we.traverse)
summary(pg)
```

Great! While it shows that `gradeRank` or `Grade` are both highly correlated... that's because it's the only
variable. We can do better. The low $R^2$ value doesn't really do much either. So let's look at a few
more variables.

Let's back up real quick and evaluate the difference between `pace` and `speed`.

While it's helpful to think in terms of pace when running, because of the math involved to transform one
metric to another, `pace` has much greater variance depending on how fast/slow I'm going.

Recall:
$$ pace = (speed[\frac{km}{hr}] * \frac{0.621 miles}{1 km} * \frac{1 hour}{60 min})^{-1} $$

If we look at both of those metrics compared side by side, Speed scales much better for lower and faster speeds.
We'll stick with using speed instead of pace, and we can transform the variable later when necessary.

```{r pace cadence, echo = F, warning = F}

g1 <- ggplot(we.traverse, 
             aes(x = Cadence, y = speed.km)) + 
  geom_point(position = "jitter", alpha = 0.3) +
  ggtitle("Speed vs Cadence") +
  labs(y = "Speed [km/h]") +
  theme_hc() +
  scale_y_continuous(breaks = seq(0, 100, 10),
                     labels = seq(0, 100, 10),
                     limits = c(0, 90))

g2 <- ggplot(we.traverse, 
             aes(x = Cadence, y = pace)) + 
  geom_point(position = "jitter", alpha = 0.3) +
  ggtitle("Pace vs Cadence") +
  labs(y = "Pace [min/mi]") +
  theme_hc() +
  scale_y_continuous(breaks = seq(0, 100, 10),
                     labels = seq(0, 100, 10),
                     limits = c(0, 90))

grid.arrange(g1, g2, ncol = 2)

rm(g1, g2, pg)
```

In going over these data, I began to think of other ways to categorize things and thought of one that may be helpful, especially
in terms of ultras -- whether I'm walking or running. An easy way to pull that out would be to look at the cadence. While this
isn't always the case, a good indicator of whether I'm running vs walking/hiking is a threshold value in cadence.

Obviously this will vary depending on the person, so I'll have a look at what my values look like.

```{r walk cadence, echo = F}
ggplot(we.traverse, aes(Cadence)) +
  geom_histogram(binwidth = 5) +
  scale_x_continuous(breaks = seq(0, 150, 5)) +
  theme_hc() +
  ggtitle("Cadence over the Traverse")
```

It would appear that I typically run at between 80-90 spm, while I walk at around 55-60 spm. Note that _spm_ is
_steps per minute_ measuring a single leg, not both. A good threshold value would be right at 70.

We can now add a cateogry to our data that tells us whether we are walking or not and then see if that makes sense in terms of
speed.

```{r walk category, echo = FALSE}
we.traverse <- we.traverse %>% 
  mutate(MovementType = ifelse(
    Cadence < 70, 1, 0)
  )
```
```{r plot walk or run}
ggplot(we.traverse, aes(x = speed.km, y = Cadence, col = factor(MovementType))) +
  geom_point() +
  theme_hc() +
  scale_color_hc(name = "Movement",
                 breaks = c(0, 1),
                 labels = c("Running", "Walking"))
```

Hmm, something seems off. Maybe it's not as simple as I thought.
Even when my cadence was in the "walking" region, my speed seemed to still
indicate I was at running speed. Perhaps, on the downhills, I was taking longer strides?

```{r vertical cadence}
ggplot(we.traverse, aes(x = Duration/3600, y = Cadence)) +
  geom_point(aes(col = factor(MovementType))) +
  theme_hc() +
  scale_color_hc(name = "Movement",
                 breaks = c(0, 1),
                 labels = c("Running", "Walking")) +
  scale_x_continuous("Duration [Hours]",
                     breaks = seq(0, 12, 1),
                     labels = seq(0, 12, 1)) +
  geom_line(aes(y = Altitude))
```

All I'm gathering from this is there is no easy distinction to determine whether I'm walking or not. Because of that,
we will remove the categorical variable for `MovementType`. It might be easier to extract than I realize, but for
now we'll remove that category and can explore it later.

```{r remove MovementType, echo = F}
we.traverse <- we.traverse %>% select(-MovementType)
```

## Developing a Model
Now let's look at a few of the correlations between the metrics recorded.

```{r speed est, echo = F}
model.sc <- lm(speed.km ~ Cadence, data = we.traverse)
summary(model.sc)

model.scA <- lm(speed.km ~ Cadence + Altitude, data = we.traverse)
summary(model.scA)

model.scHA <- lm(speed.km ~ Cadence + Altitude + HeartRate, data = we.traverse)
summary(model.scHA)

model.scHG <- lm(speed.km ~ Cadence + Grade + HeartRate, data = we.traverse)
summary(model.scHG)

model.scHGA <- lm(speed.km ~ Cadence + Altitude + HeartRate + Grade, data = we.traverse)
summary(model.scHGA)


## Clear some memory
rm(model.sc, model.scA, model.scHA, model.scHG, model.scHGA)
```

It may take a little to look through the different models, but we're just comparing the different variables and they're output effect.

Something else that may be quite helpful is the duration to which the event has taken. For example, look at the
`gradeRank vs Pace` graph, and around the 0% - 2.5% grade region, there are some high points, but also some low points
that are in the 'walking' range (lower).

2.5% is a pretty mild grade. However, depending on the time of day and how my legs were feeling, I may have decided
to walk even a small hill instead. It may not always be an influencing factor, but I think it may shed some additional
light on the entire picture.

Let's add duration to our model.

```{r speed est duration, echo =F}

model.scHGAD <- lm(speed.km ~ Cadence + Altitude + HeartRate + Grade + Duration, data = we.traverse)
summary(model.scHGAD)

model.scHGADT <- lm(speed.km ~ Cadence + Altitude + HeartRate + Grade + Duration + Temperature, data = we.traverse)
summary(model.scHGADT)

rm(model.scHGADT, model.scHGAD)
```

This part is still under exploration and am not sure how to proceed here, but will come back to it once I learn more.