---
title: "Logistic Regression - NHIS"
author: "Tomas"
date: "August 28, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

## Regression with Binary Outcomes

### Logistic Regression
***

The text below is adapted from documents provided from Harvard University's statistical software
workshop along with added problems from Springboard.

This far we have used the `lm` function to fit our regression models.
`lm` is great, but limited-in particular it only fits models for
continuous dependent variables. For categorical dependent variables we can use the `glm()' function.

For these models we will use a different dataset, drawn from the 
National Health Interview Survey. From the CDC website:

> The National Health Interview Survey (NHIS) has monitored
> the health of the nation since 1957. NHIS data on a broad
> range of health topics are collected through personal
> household interviews. For over 50 years, the U.S. Census
> Bureau has been the data collection agent for the National
> Health Interview Survey. Survey results have been
> instrumental in providing data to track health status,
> health care access, and progress toward achieving national
> health objectives.

Load the National Health Interview Survey Data [^1]:
```{r load NHIS data, message=FALSE}
setwd("C:/Users/castillo/OneDrive/School/Springboard/FoundationsDS/in-depth/LogisticProblemSet/")
NH11 <- readRDS("dataSets/NatHealth2011.rds")
labs <- attributes(NH11)$labels
```

## Logistic Regression Example

Let's predict the probability of being diagnosed with hypertension based on `age_p`, `sex`, `sleep`, and `bmi`.
First, we'll check the structure of hypertension (`hypev`) and reformat the factoring levels.

```{r refactor hypertension}
str(NH11$hypev) # Check the structure

levels(NH11$hypev) # Check the leveles

summary(NH11$hypev) # Current state - note the NAs

## Collapse all missing values to NA
NH11$hypev <- factor(NH11$hypev, levels = c("2 No", "1 Yes"))
summary(NH11$hypev) # Updated output
```

With the factors adjusted, we'll now create our initial regression model.

```{r Initial regression model}
hyp.out <- glm(hypev ~ age_p + sex + sleep + bmi,
               data = NH11,
               family = 'binomial')

coef(summary(hyp.out))
```

### Logistic Regression Coefficients

The generalized linear models (`glm`) use link functions, so raw coefficients are difficult
to interpret. For example, the age coefficient of 0.06 in the previous model tells us that for every
one unit increase in age, the log odds of hypertension diagnosis increases by 0.06.
Since most of us are not used to thinking in log odds, this is not too helpful!

One solution is to transform the coeffcients to make them easier to interpret:

```{r initial regression transformation}
hyp.out.tab <- coef(summary(hyp.out))
hyp.out.tab[, 'Estimate'] <- exp(coef(hyp.out))
hyp.out.tab
```

### Generating Predicted Values

In addition to transforming the log-odds produced by `glm` to odds, we can use the
`predict()` function to make direct statements about the predictor in our model. For example, we can ask
"How much more likely is a 63 year old female to have hypertension compared to a 33 year old female?".

To answer this question, we'll create a custom dataset that has the predictors set at specific levels.

```{r Predicting levels}
## Createing the dataset from our full NH11
predData <- with(NH11,
                 expand.grid(age_p = c(33, 63),
                             sex = "2 Female",
                             bmi = mean(bmi, na.rm = TRUE),
                             sleep = mean(sleep, na.rm = TRUE)))

## Predict hypertension at those levels
cbind(predData, predict(hyp.out,
                        type = 'response',
                        se.fit = TRUE,
                        interval = 'confidence',
                        newdata = predData))
```

This tells us that a 33 year old female has a 13% probability of
having been diagnosed with hypertension, while a 63 year old female has a 48% probability of having been
diagnosed.

## Packages for Computing and Graphing Predicted Values
***

Instead of doing all this ourselves, we use the `effects` package to compute quantities of interest
for us (cf. the Zelig package).

```{r effects package, warning=FALSE}
library(effects)
plot(allEffects(hyp.out))
```

## Exercise: Logistic Regression
***

Now it's time to give this a try for ourselves.

Using the NH11 data set from above:

1. Use `glm` to conduct a logistic regression to predict ever worked (`everwrk`) using age (`age_p`)
and marital status (`r_maritl`).

```{r ever worked initial}
summary(NH11$everwrk)

wrkd.mod1 <- glm(everwrk ~ age_p + r_maritl, 
            data = NH11,
            family = 'binomial')

summary(wrkd.mod1)
```

There are quite a few NAs and some other non-yes/no. We are really only interested in the Yes/No responses. But if we
were to exclude non-Yes/No's, that still leaves us with 12153 Yes's and 1887 No's. Since `everwrk` is our dependent variable,
we can't really make up values for it, and since each observation represents a single person's response, it's best if
we eliminate that person from our dataset for this model.

The other variable that we'll look at is `age_p`, and `r_maritl`.
```{r summary of age and marital status}
summary(NH11$age_p)
hist(NH11$age_p, breaks = seq(18, 85, 1))

summary(subset(NH11, !is.na(everwrk))$r_maritl)
levels(NH11$r_maritl)
plot(NH11$r_maritl,
     legend.text = levels(NH11$r_maritl))

NH11 %>% group_by(r_maritl) %>% summarise(count = n())

```

By looking at the variables we're going to be using for our model, we'll make a few adjustments:

1. Remove all NA observations of `everwrk`.

2. Filter the ages to exclude 18 and 85-year olds since those numbers are most likely erroneous points. It's a survey, so probably has to do
with the surveyor not getting an age but determining the subject was young (18) or old (85) (just an assumption).

3. We are only concerned with the binary response of `everwrk`: that is `1 Yes` or `2 No`.

4. Include only the categories of marital status with a response (there are 8).

```{r subset of NH11}
NH11.sub <- subset(NH11, age_p > 18 & age_p < 85 & !is.na(everwrk) & 
                     (everwrk == "1 Yes" | everwrk == "2 No"))
```

Now that we have that set up - let's refine our model and see if it's improved.

```{r correlation matrix}
wrkd.mod2 <- glm(everwrk ~ age_p + r_maritl,
                 data = NH11.sub,
                 family = "binomial")

summary(wrkd.mod2)

```

The first model gave us an AIC of 10,429; while the second one has a value of 9406. Not a huge improvement, but better.


2. Predict the probability of working for each level of marital status.

We can use the `predict` function to give us probabilities of whether or not someone was working, regardless of `age_p`.

```{r predicting}
## Setup variable for proper r_maritl levels (non-zero)
usable.maritl <- levels(NH11.sub$r_maritl)[
  which(table(NH11.sub$r_maritl) > 0)]

## Setup a predicting
predData <- with(NH11.sub,
                 expand.grid(r_maritl = usable.maritl,
                             age_p = mean(bmi, na.rm = TRUE)))

cbind(predData,
      predict(wrkd.mod2,
              type='response',
              se.fit=TRUE,
              interval = 'confidence',
              newdata = predData))

```

What does this tell us? Since we are holding the age constant at the average for the entire dataset, we are only
varying the marital status. When looking at the graph of the populated marital status, the most popular category
is "Married - spouse in household". Translation - if you are married and have a spouse in the household, there is
a 20% that you have worked. Whereas if you are widowed, there is a 30% chance you have worked.

We can look at our independent variable's effect on whether or not an individual has worked through use of the `effects` plot.

```{r effect of working}
plot(allEffects(wrkd.mod2))
```

While not the most useful of information, or best prediction, it's at least a good exercise into determinig what the
data tells us.


[^1]: [CDC Website](http://www.cdc.gov/nchs/nhis.html)
