---
title: "Refining Data"
author: "Tomás"
date: "March 7, 2016"
output: html_document
---
```{r, echo=FALSE, message=FALSE}
library(dplyr)
library(tidyr)

setwd("C:/Users/castillo/OneDrive/School/SlideRule/FoundationsDS/dataWrangling/")
df_original <- tbl_df(read.csv("refine_original.csv", stringsAsFactors = F))
```
## Overview
Section 3.1 of Data Wrangling goes over the fundamentals of how to refine data. The original data given looks like below:

```{r echo=FALSE}
head(df_original)
```

### Typos And Company Names

There are misspellings (e.g. '`r paste(df_original$company[1])`' which should be 'philips') and a
column that combines a product code with its product number (e.g. '`r paste(df_original$Product.code...number[1])`').

I made a dictionary type list that contained the different first letters of the companies, and the standardized name to go with.

```{r}
comp_char <- list(c("f", "a","v","u"), 
                  c("philips","akzo", "van houten", "unilever"))
```


The list made it easy to run a _for_ loop over different companies. After analyzing the types of typos, I made sure to iterate over the 'philips' first
since it had a first letter typo. I created a _len_ column that displayed the company's number of characters, but ended up not needing it.
With my filters in place, I created a temporary dataframe (storing only the companies I wanted) and added the rows together using rbind. I wanted to
practice using multiple methods of combining data.

```{r}
for (i in 1:4) {
  if (i == 1) {
    
    df_original <- df_original %>% 
      mutate(company = tolower(company), 
             len = nchar(company), 
             first_char = substr(company, 1, 1))
    
    df <- df_original %>% 
      filter(first_char == "p" | first_char == "f") %>% 
      mutate(company = comp_char[[2]][i])
    
  } else {
    
    df_temp <- df_original %>% 
      filter(first_char == comp_char[[1]][i]) %>% 
      mutate(company = comp_char[[2]][i])
    
    df <- rbind(df, df_temp)
  }
}

## Clean up a little
rm(df_temp)
```

### Products & Categories

We can use the built-in `TIDYR::separate` to split up the columns and create a list similar to before for the categories of the product code.
This time, I used a little more efficient method of creating the columns.

Also included the step to add the full address and re-order columns using `DPLYR::select(everything())`.

```{r}
prod_cat <- list(p = 'smartphone', v = 'tv', x = 'laptop', q = 'tablet')

df <- df %>% separate(Product.code...number, 
                      c('product_code', 'product_number'), sep='-')

df <- df %>% mutate(product_category = paste0(prod_cat[product_code]), 
                    full_address = paste(address, city, country, sep = ","))

df <- df %>% select(-(address:name),everything())

## Clean up a little more
df$len <- NULL
df$first_char <- NULL
```

### Binary Columns

When adding the binary columns, we're basically creating a flag for companies.

```{r}
df$company == 'philips'
```

An easy solution was to just convert the the `TRUE | FALSE` to `1 | 0`. Since I already had a list of the product categories, and company names, it made this step relatively quick with a _for_ loop.

```{r}
### Since we already have a list defined for the companies, we'll use a for loop to go through each one and
###  evaluate each company, while adding a header that corresponds to the company we're evaluating

for (i in 1:4) {
  i_company <- comp_char[[2]][i]
  i_header <- paste("company", i_company, sep = "_")
  df[[paste0(i_header)]] <- as.numeric(df$company == paste0(i_company))
}

### Similarly, we'll do the same thing with the product category
### We'll just use the product's code (p, v, x, q) along with it's list we created earlier

for (i in 1:4) {
  i_product <- names(prod_cat)[i]
  i_header <- paste("product", prod_cat[i], sep = "_")
  df[[paste0(i_header)]] <- as.numeric(df$product_code == paste0(i_product))
}

### And clean up the mess again
rm(i_company, i_header, i_product)
```

### Output

Lastly, we'll write the output file and consider this done!

```{r}
head(tbl_df(df))
write.csv(df, file = "refine_clean.csv")
```
